---
title: "Lab 6: Childcare Costs in California"
author: "Alondra Casas"
format: gfm
editor: source
execute: 
  echo: true
---

# Part One: Set-up

In this lab, we will be using the **tidyr** and **forcats** packages to explore the cost of childcare across the US. **You are expected to use functions from tidyr and forcats to do your data manipulation!**

<!-- See instructions for words of advice on completing the assignment! -->

# Exploring Childcare Costs

## The Data

In this lab we're going look at the median weekly cost of childcare in California. The data come to us from [TidyTuesday](https://github.com/rfordatascience/tidytuesday). A detailed description of the data can be found [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-09/readme.md). **You will need to use this data dictionary to complete the lab!**

We also have information from the California State Controller on tax revenue for california counties from 2005 - 2018. I compiled the data from [this website](https://counties.bythenumbers.sco.ca.gov/#!/year/default) for you. Note that there is no data for San Franscisco County. The variables included in the `ca_tax_revenue.csv` data file (loaded below) include:

-   `entity_name`: County name
-   `year`: fiscal year
-   `total_property_taxes`: total revenue in \$ from property taxes
-   `sales_and_use_taxes`: total revenue in \$ from sales and use taxes

**0. Load the appropriate libraries and the data.**

```{r}
#| label: packages
library("tidyverse")
library(forcats)
```

```{r}
#| label: load-data

childcare_costs <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/childcare_costs.csv')

counties <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/counties.csv')

tax_rev <- read_csv('https://raw.githubusercontent.com/statistical-computing-r/spring-2026/refs/heads/main/labs/instructions/data/ca_tax_revenue.csv')
```

**1. Briefly describe the `childcare_costs` dataset (\~ 4 sentences). What information does it contain?**

##### The childcare costs dataset contains information about the country, year, and proportions for a lot of factors such as race and population. Some of the columns have confusing names so it's a little hard to decipher what they mean but I think they are all proportional values for different hosuehold factors. It also has a lot of different columns, with a lot of them containing zeros. I also thougt it was interesting how the countries were labeled by code.

## California Childcare Costs

**2. Let's start by focusing only on California. Create a `ca_childcare` dataset of childcare costs in California, containing (1) county information and (2) all information from the `childcare_costs` dataset. You should do all of this within one pipeline.**

```{r}
#| label: ca-childcare-costs
ca_childcare <- 
  counties |>
  filter(state_name == "California") |>
  inner_join(childcare_costs)
```

<!-- Checkpoint: There are 58 counties in CA and 11 years in the dataset. Therefore, your new dataset should have 53 x 11 = 638 observations. -->

**3. Now, lets add the tax revenue information to the `ca_childcare` dataset. Add the data from `tax_rev` for the counties and years that are already in the `ca_childcare` data. Overwrite the old `ca_childcare` data with this dataset.**

```{r}
#| label: add-tax-information

ca_tax <- ca_childcare |>
  left_join(tax_rev, by = join_by(county_name == entity_name,
                                  study_year == year))

```

<!-- Checkpoint: You are only adding columns here, so your new dataset should still have 638 observations! -->

**4. Using a function from the `forcats` package, complete the code below to create a new variable where each county is categorized into one of the [ten (10) Census regions](https://census.ca.gov/regions/) in California. Use the Region description (from the plot), not the Region number.** The code below will help you get started.

```{r}
#| label: defining-census-regions

superior_counties <- c("Butte","Colusa","El Dorado",
                       "Glenn","Lassen","Modoc",
                       "Nevada","Placer","Plumas",
                       "Sacramento","Shasta","Sierra","Siskiyou",
                       "Sutter","Tehama","Yolo","Yuba")

north_coast_counties <- c("Del Norte","Humboldt","Lake",
                          "Mendocino","Napa","Sonoma","Trinity")

san_fran_counties <- c("Alameda","Contra Costa","Marin",
                       "San Francisco","San Mateo","Santa Clara",
                       "Solano")

n_san_joaquin_counties <- c("Alpine","Amador","Calaveras","Madera",
                            "Mariposa","Merced","Mono","San Joaquin",
                            "Stanislaus","Tuolumne")

central_coast_counties <- c("Monterey","San Benito","San Luis Obispo",
                            "Santa Barbara","Santa Cruz","Ventura")

s_san_joaquin_counties <- c("Fresno","Inyo","Kern","Kings","Tulare")

inland_counties <- c("Riverside","San Bernardino")

la_county <- "Los Angeles"

orange_county  <- "Orange"

san_diego_imperial_counties <- c("Imperial","San Diego")
```

```{r}
#| label: recoding-county-to-census-regions
# Finish this code using the census regions defined above

ca_childcare <- ca_childcare |> 
  mutate(county_name = str_remove(county_name, " County")) |>
    mutate(region = fct_collapse(county_name,
    'Superior California' = c("Butte", "Colusa", "El Dorado","Glenn","Lassen","Modoc",
                       "Nevada","Placer","Plumas",
                       "Sacramento","Shasta","Sierra","Siskiyou",
                       "Sutter","Tehama","Yolo","Yuba"),
    'North Coast' = c("Del Norte","Humboldt","Lake",
                          "Mendocino","Napa","Sonoma","Trinity"),
    'San Francisco' = c("Alameda","Contra Costa","Marin",
                       "San Francisco","San Mateo","Santa Clara",
                       "Solano"),
    'Northen San Joaquin' = c("Alpine","Amador","Calaveras","Madera",
                            "Mariposa","Merced","Mono","San Joaquin",
                            "Stanislaus","Tuolumne"),
    'Central Coast' = c("Monterey","San Benito","San Luis Obispo",
                            "Santa Barbara","Santa Cruz","Ventura"),
    'South San Joaquin' = c("Fresno","Inyo","Kern","Kings","Tulare"),
    'Inland' = c("Riverside","San Bernardino"),
    'Los Angeles' = "Los Angeles",
    'Orange County' = "Orange",
    'San Diego' = c("Imperial","San Diego")
    
  ))
  
```

<!-- Tip: I have provided you with code that eliminates the word "County" from each of the county names in your `ca_childcare` dataset. You should keep this line of code and pipe into the rest of your data manipulations. -->

**5. Let's consider the median household income of each region, and how that income has changed over time. Create a table with ten rows, one for each region, and two columns, one for 2008 and one for 2018 (plus a column for region). The cells should contain the `median()` of the median household income (expressed in 2018 dollars) of the `region` and the `study_year`. Order the rows by 2018 values from highest income to lowest income.**

<!-- Tip: This will require transforming your data! Sketch out what you want the data to look like before you begin to code. You should be starting with your California dataset that contains the regions! -->

```{r}
#| label: median-income-by-region-over-time

ca_childcare |>
  filter(study_year %in% c(2008, 2018)) |> 
  group_by(region, study_year) |>
  summarise(medianincome = median(mhi_2018, na.rm = TRUE), .groups = "drop") |>
      pivot_wider(names_from = study_year, values_from = medianincome) |>
        arrange(desc('2018'))

```

**6. Which California `region` had the lowest `median` full-time median weekly price for center-based childcare for infants in 2018? Does this `region` correspond to the `region` with the lowest `median` income in 2018 that you found in Q4?**

```{r}
#| label: lowest-median-weekly-price-2018

ca_childcare |>
  filter(study_year == 2018) |>
  group_by(region) |> 
  summarise(infant_care = median(mc_infant, na.rm = TRUE)) |>
  arrange(infant_care) |>
  slice(1)


```

<!-- Checkpoint: The code should give me the EXACT answer. This means having the code output the exact row(s) and variable(s) necessary for providing the solution. -->

**7. The plot in the instructions shows, for all ten regions, the change over time of the full-time median price for center-based childcare for infants, toddlers, and preschoolers. Recreate the plot. You do not have to replicate the exact colors or theme, but your plot should have the same content, including the order of the facets and legend, reader-friendly labels, axes breaks, and a loess smoother.**

<!-- See instructions for hints on recreating the plot! -->

```{r}
#| label: recreate-plot

ca_childcare |>
  ggplot(aes(x = study_year, 
                   y = median_price,
                   color = region)) +
  geom_point(alpha = 1) 
```
